Outline flowCore paper

### 2008-04-17 paper/outline.tex
more on the outline
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31140 bc3139a8-67e5-0310-9ffc-ced21a209358

\subsection{R and Bioconductor}
R is a solid statistical programing environment, particularly geared
towards the analysis of large numeric datasets typical to arise from
high throughput experiments such as microarray gene expression analysis
or FC-HCS. Today, the R open source project offers a wide range of
statistical and visualization methods developed for various fields of
application. In addition, R is a research platform which
bioinformaticians, computer scientists, and statisticians routinely
use to develop novel tools and methodology for statistical data
analysis. For biological applications, the Bioconductor project has
become the standard toolset \citep{gentleman2006bos} to process data
spanning a diversity of research fields from genomics to proteomics to
complex cell biology. This shared platform proved to be especially
beneficial for the development of analysis routines for microarray gene
expression data, and by now many of these innovations have been
incorporated into commercial software products, thus greatly improving
the quality of gene expression data analysis. Shared data structures
have been shown to be essential for this kind of collaborative effort.

\subsection{Existing data standards and conventions}
Currently, data from flow cytometry experiments are stored in single
files following a well-defined data standard (FCS). This standard
tries to capture as much of the accompanying meta information as
possible, most of which is produced by the measurement instrument or
has to be supplied manually by the operator upon experiment
setup. However, the recent developments in high-throughput flow
cytometry are shifting the focus of interest away from single-tube
based measurements towards large and complex experimental designs with
dozens of co-variates and influencing factors. Self-contained data
structures are a prerequisite to allow for coordinated analysis of
such data sets. Most of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for the mostly GUI-driven user interfaces and hence are
not easily amenable to programmatical access. In addition, the
closed-source nature of most of these products makes it virtually
impossible to connect and integrated these data.

Traditionally, the majority of flow cytometry experiments have been
analysed manually by direct data inspection or very basic comparisons
of summary statistics. We believe that these approaches do not give
justice to the highly complex nature of FCM data, essentially
discarding many of the fundamental aspects such as data distribution
or high-dimensionality. In addition, the subjective character of
manual analyses are a major obstacle for reproducibility. For FC-HCS
data, unassisted manual inspection is no longer feasible, and robust
statistical methods need to be developed to point the investigator to
the interesting aspects of the data or to potential problems. While the
expert knowledge of immunologists and researchers remains crucial for
the understanding of FCM data, we believe that tight collaboration
with other research fields such as statistics and computer science can
greatly improve the relevance of flow cytometry in today's
high-throughput biological sciences.

@@

@@

The flowCore framework presented here offers import functionality for
raw data FCS files along with a complete set of file-specific meta information.
Moreover, it is one of the very first software implementations of the Gating-ML CR, 
which opens the possibility to integrate flowCore in existing work flows and to 
communicate with virtually every other FCM tool that adheres to this standard.

{\bf NLM:} Only a couple of commercial software (e.g., FlowJo and
DeNovo) allows the user to analyse batch of FCS files.  Only a few
allows export in XML (e.g., FlowJo). Most commercial software require
a Microsoft Windows platform whereas flowCore runs on multiple
operating systems including Windows, Mac OS X and Unix.

@@

\section{Basic data structures}
\subsection{flowFrame}
flowCore's primary task is the representation and basic manipulation
of flow cytometry data. This is accomplished through a
data model very close to that adopted by other Bioconductor packages
used for gene expression analysis and thus familiar to most
Bioconductor users. Similar to any other flow tool, the basic unit of
organization in flowCore is a collection of events. We call the
structure that hold this data a flowFrame. All relevant information
for the collection are stored along with the matrix of raw data values
and can be accessed programmatically. Most prominently, these are
descriptors of the stains used in the experiment and the respective
measurement channels, information about compensation performed at the
instrument side and any additional keywords the user deems to be
important to annotate the data. A number of quality checks are
performed during the creation of a flowFrame to ensure data
integrity. There is an extended interface that allows for the
standardized interaction with the flow data, thus enabling the
development of more specialized tools for specific tasks of the data
analysis process.

\subsection{flowSet}
In high-throughput flow cytometry, many of the analysis tasks need to
be performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flowFrames called a
flowSet. flowSets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flowFrames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of mixups and errors that are
all too common when dealing with complex data in the ususal way of
more or less ad hoc spreadsheets that lack a controlled structure. In
general, there is no native ordering to a flowSet, which in a way
reflects the still predominant application of tube-based experiments
rather than experiments following a plate-design. However, the flowSet
structure can be readily extended to also accommodate such spacial
information \textit{(reference)}

{\bf NLM:} How about mentioning PlateCore here?

@@

{\bf PH:} I think the Frame/Set concept is pretty interesting and
powerful. Was it originated for this project? The ability to program
sets flowCore aside from all other flow software applications, I
think. I think there is something really important and valuable about
how the data structures are ogranized. I'm not sure how to write that
in a way to appeal to Cytometry A readers, but I think we could work
this out.

\section{Standard flow operations}
The basic operations in flow cytometry analysis are almost always the
same: the data need to be compensated (if not already done on the
instrument side), often transformed and sup-population of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for flow data analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven spirit. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flowFrames as well as on entire
flowSets. While transformation and to a certain extent also
compensation are fairly routine operations with only limited potential
for improvement, being able to implement new methodology for gating of
flow cytometry data and extending the capabilities of flowCore through
the usual ways of object oriented programming is a feature, that
clearly sets our framework apart from all the other flow toolsets
around. By factoring out as much of the bookkeeping as possible,
authors can focus on the actual operations rather then having to deal
with the tedious details of data integration and access. Third party
methods implemented by others can act on their own as first-class
citizens in the analysis, without breaking the workflow or the basic
infrastructure. This design allows for the straight forward extension
of flowCore's capabilities and already fostered the development of a
number of valuable improvements. \textit{(cite stuff here)}

\subsection{transformations}
All the major transformations that are routinely used in flow
cytometry analysis have been implemented in flowCore. Furthermore, the
design of the R language makes it easy to define arbitrary functions
to apply to the data of individual flowFrames or entrire flowSets,
respectively. Proper transformation have been proven beneficial for
both data visualization and modelling \textit{cite flowClust paper}.

\subsection{gating}
In flowCore, gates are represented as individual objects that can be
extended in an object oriented manner. All basic gate types such as
rectangular gates, ellipses and polygons are implemented as part of
the base framework. In addition, we introduce the notion of
data-driven gates, for which the necessary parameters are computed
based on the properties of the underlying data, for instance
distribution or density estimates. The ability to programmatically
access gates is a prerequisite for automated or semi-automated
gating. By utilizing a unified interface for all different types of
gates, the user is able to subset data sets as well as create summary
statistics, for instance of proportions of events falling in single
gates or combinations of multiple gates. The definition of gates in
flowCore follows the Gating-ML CR, thus any flowCore
gating strategy can be reproduced by any other software that also
adheres to the standard.

\section{Related flow packages}
In addition to the flowCore package that offers basic infrastructure
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of flow cytometry data
analysis. The flowViz package provides sophisticated data
visualization tools that make use of multivariate trellis
plotting. These functions can be used to quickly generate customized
plots for extended cytometry data sets for both direct data inspection
and quality control. The flowQ package offers more advanced quality
assurance (QA) methodology and a framework to create interactive
reports of QA results. Most of the flowUtils content deals with data
import and output from and to flowCore and the flow-cytometry specific
standard markup languages. Finally, flowStats provides statistical
methods that are relevant in the context of flow cytometry data
analysis.

>>>>

\subsection*{R and Bioconductor}
R is a solid statistical programing environment, particularly geared
towards the analysis of large numeric datasets typical to arise from
high throughput experiments such as microarray gene expression
analysis or FC-HCS. Today, the R open source project offers a wide
range of statistical and visualization methods developed for various
fields of application. In addition, R is a research platform which
bioinformaticians, computer scientists, and statisticians routinely
use to develop novel tools and methodology for statistical data
analysis. For biological applications, the Bioconductor project has
become the standard toolset \citep{gentleman2006bos} to process data
spanning a diversity of research fields from genomics to proteomics to
complex cell biology. This shared platform proved to be particularly
beneficial for the development of analysis routines for microarray
gene expression data, and by today many of these early innovations
have been incorporated into commercial software products, thus greatly
improving the quality of gene expression data analysis. In this
context, shared data structures have been shown to be essential for
such a collaborative effort.

\subsection*{Existing data standards and conventions}
Currently, data from flow cytometry experiments are stored in single
files following a well-defined data standard (FCS). This standard
attempts to capture as much of the accompanying meta information as
possible, most of which is produced by the measurement instrument or
has to be supplied manually by the operator upon experiment
setup. However, the recent developments in high-throughput flow
cytometry are shifting the focus of interest away from single-tube
based measurements towards large and complex experimental designs with
dozens of co-variates and influencing factors. Self-contained and
self-reflective data structures are a prerequisite to allow for
coordinated analysis of such experiments. Most of the currently
available software solutions offer only limited support for such
structures, or make use of containers like XML or binary storage
formats that are designed specifically for the mostly GUI-driven user
interfaces and hence are not easily amenable to programmatical
access. In addition, the closed-source nature of most of these
products makes it virtually impossible to connect and integrated the
data.

Traditionally, the majority of flow cytometry experiments have been
analysed manually by direct data inspection in one ore tow dimensions,
or by very basic comparisons of summary statistics. We believe that
these approaches do not give justice to the highly complex nature of
FCM data, in particualar, they discard many of the fundamental aspects
such as underlying distribution of the data or the
high-dimensionality. In addition, the subjective character of manual
analyses are a major obstacle for reproducibility. For FC-HCS data,
unassisted manual inspection is no longer feasible, and robust
statistical methods need to be developed to point the investigator to
the interesting aspects of the data or to potential problems. While
the expert knowledge of immunologists and researchers remains crucial
for the understanding of FCM data, we believe that tight collaboration
with other research fields such as statistics and computer science can
greatly improve the relevance of flow cytometry in today's
high-throughput biological sciences.

@@

The flowCore framework presented here offers import functionality for
raw data FCS files along with their complete set of file-specific meta
information.  Moreover, it is one of the very first software
implementations of the Gating-ML CR, which opens the possibility to
integrate flowCore in existing work flows and to communicate with
virtually every other FCM tool that adheres to the proposed
standard. In addition, the software is platform independent, running
on various operating systems including Windows, Mac OS X and
Linux/Unix.

flowCore is not a GUI-driven software, and all operations are
currently done using a command line interface. Conceptually, it is
possible to add a more elaborate user interface on top of that,
however we strongly focus on programmatic access to enable the
convenient development of novel analysis methods. By taking the burden
of data management from the programmer and by providing well-defined
APIs and a structured class hierarchy, it is possible to readily test
new ideas and to easily extend the frameworks functionality. Both the
underlying R and Bioconductor software and flowCore itself are open
source and open development projects, thus community involvement in the
developing process is highly appreciated.

@@

{\bf NLM:} Only a couple of commercial software (e.g., FlowJo and
DeNovo) allows the user to analyse batch of FCS files.  Only a few
allows export in XML (e.g., FlowJo).

@@

\section{Basic data structures}
\subsection*{flowFrame}
flowCore's primary task is the representation and basic manipulation
of flow cytometry data. This is accomplished through a data model very
close to that adopted by other Bioconductor packages used for gene
expression analysis, and thus familiar to most Bioconductor
users. Similar to any other flow tool, the basic unit of organization
in flowCore is a collection of events. We call the structure that hold
this data a flowFrame. All relevant information for the collection are
stored along with the matrix of raw data values and can be accessed
programmatically. Most prominently, these are descriptors of the
stains used in the experiment and the respective measurement channels,
information about compensation performed at the instrument side and
any additional keywords the user deems to be important to annotate the
data. A number of quality checks are performed during the creation of
a flowFrame to ensure data integrity. There is an extended interface
that allows for the standardized interaction with the flow data, thus
enabling the development of more specialized tools for specific tasks
of the data analysis process.

\subsection*{flowSet}
In high-throughput flow cytometry, many of the analysis tasks need to
be performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flowFrames called a
flowSet. flowSets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flowFrames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of mixups and errors that are
all too common when dealing with complex data in the ususal way:
namely, more or less ad hoc spreadsheets that lack a controlled
structure. In general, there is no native ordering to a flowSet, which
in a way reflects the still predominant application of tube-based
experiments rather than experiments following a plate-design. However,
the flowSet structure can be readily extended to also accommodate such
spatial information \citep{(plateCore)}

@@

\section{Standard flow operations}
The basic operations in flow cytometry analysis are almost always the
same: the data need to be compensated (if not already done on the
instrument side), often transformed and sup-population of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for flow data analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven spirit. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flowFrames as well as on entire
flowSets. While transformation and to a certain extent also
compensation are fairly routine operations with only limited potential
for improvement, being able to implement new methodology for gating of
flow cytometry data and extending the capabilities of flowCore through
the usual ways of object oriented programming is a feature, that
clearly sets our framework apart from all the other flow toolsets
around. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and access. Third
party methods implemented by others can act on their own as
first-class citizens in the analysis framework, without breaking the
workflow or the basic infrastructure. This design allows for the
straight forward extension of flowCore's capabilities and already
fostered the development of a number of valuable
improvements. \citep{(flowClust, plateCore)}

\subsection*{Transformation and compensation}
Proper transformation have been proven beneficial for both data
visualization and modelling \textit{cite flowClust paper}. All the
major transformations that are routinely used in flow cytometry
analysis have been implemented in flowCore. Furthermore, the design of
the R language makes it easy to define arbitrary functions to apply to
the data of individual flowFrames or entrire flowSets,
respectively. Compensation is available for both flowFrames and
flowSets, and the software also offers functionality to compute
spillover or compensation matrices from a set of appropriate
compensation samples.

\subsection*{Gating}
In flowCore, gates are represented as individual classes that can be
extended in an object oriented manner. All basic gate types such as
rectangular gates, ellipses and polygon gates are implemented as part
of the base framework. In addition, we introduce the notion of
data-driven gates, or filters, for which the necessary parameters are
computed based on the properties of the underlying data, for instance
by modelling underlying data distribution or by density
estimation. The ability to programmatically access gates is a
prerequisite for automated or semi-automated gating. By utilizing a
unified interface for all different types of gates, the user is able
to subset data sets as well as to create summary statistics, like the
proportions of events falling in single gates or in the combinations
of multiple gates. Conmplex combinations and hierarchies of gates can
be captured in objects of class filterSet, allowing to apply
multi-step gating strategies. The definition of gates in flowCore
follows the Gating-ML CR, thus any flowCore gating strategy can be
reproduced by any other software that also adheres to the standard.

\section{Related flow packages}
In addition to the flowCore package that offers basic infrastructure,
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of flow cytometry data
analysis. The flowViz package provides sophisticated data
visualization tools, that make use of multivariate trellis
plotting. These functions can be used to quickly generate customized
plots for extended cytometry data sets for both direct data inspection
and quality control. The flowQ package offers more advanced quality
assurance (QA) methodology and a framework to create interactive
web-based reports of QA results. Most of flowUtil's content deals with
data import and output from and to flowCore and the flow-cytometry
specific standard markup languages. Finally, flowStats provides
statistical methods that are relevant in the context of flow cytometry
data analysis.
